Industry Internship -- Final Report

### 2023-03-21 tex/report_uds/report.tex
Added base for uds internship report

\usepackage{cleveref}
\usepackage{doc}

% use this if you have a long article and want to create an index
% \usepackage{makeidx}

\lstset{
  frame=none,
  xleftmargin=2pt,
  stepnumber=1,
  numbers=left,
  numbersep=5pt,
  numberstyle=\ttfamily\tiny\color[gray]{0.3},
  belowcaptionskip=\bigskipamount,
  escapeinside={*'}{'*},
  tabsize=2,
  emphstyle={\bf},
  commentstyle=\color{ForestGreen},
  stringstyle=\mdseries\rmfamily,
  showspaces=false,
  keywordstyle=\bfseries\rmfamily,
  columns=flexible,
  basicstyle=\small\sffamily,
  showstringspaces=false,
  morecomment=[l]\%,
}
\lstset{basicstyle=\footnotesize\ttfamily,breaklines=true}
% In order to save space or manage large tables or figures in a
% landcape-like text, you can use the rotating and pdflscape
% packages. Uncomment the desired from the below.
%
% \usepackage{rotating}
% \usepackage{pdflscape}

% Some of our commands for this guide.
%
\authorrunning{Abelt}

\newcommand{\easychair}{\textsf{easychair}}
\newcommand{\miktex}{MiK{\TeX}}
\newcommand{\texniccenter}{{\TeX}nicCenter}
\newcommand{\makefile}{\texttt{Makefile}}
\newcommand{\latexeditor}{LEd}

\newcommand{\LayeredTypes}{\textsc{LayeredTypes}}
%\makeindex

%% Front Matter
%%
% Regular title as in the article class.
%
\title{CAMELOT Research Scholarship -- Final Report}

@@

% Institutes for affiliations are also joined by \and,
\institute{
  LASIGE,
  Faculdade de Ciências da Universidade de Lisboa, Portugal\\
  \email{labelt@lasige.di.fc.ul.pt},
  \email{alcides@ciencias.ulisboa.pt}
\and
   Saarland University,
   Germany\\
}

%  \authorrunning{} has to be set for the shorter version of the authors' names;
% otherwise a warning will be rendered in the running heads. When processed by
% EasyChair, this command is mandatory: a document without \authorrunning
% will be rejected by EasyChair

% \titlerunning{} has to be set to either the main title or its shorter
% version for the running heads. When processed by
% EasyChair, this command is mandatory: a document without \titlerunning
% will be rejected by EasyChair
\titlerunning{Project Report CAMELOT}

@@

% The table of contents below is added for your convenience. Please do not use
% the table of contents if you are preparing your paper for publication in the
% EPiC Series or Kalpa Publications series

%\setcounter{tocdepth}{2}
%{\small
%\tableofcontents}

%\section{To mention}
%
%Processing in EasyChair - number of pages.
%
%Examples of how EasyChair processes papers. Caveats (replacement of EC
%class, errors).

%------------------------------------------------------------------------------
\section{Introduction}

This document is the final project report for the research scholarship granted in the scope of the CAMELOT project for the duration of 01.09.2022-30.11.2022. It was carried out by Lukas Abelt in the LASIGE Research Group within the Reliable Software Systems Research Line. The work was supercised by Alcides Fonseca, Assistant Professor at the Faculdade de Ciências da Universidade de Lisboa. The work was part of an Erasmus internship that ends in March 2023.

\section{Description of the Scholarship}
\subsection{Context}
\label{sec:context}
Most programmers and computer scientists will be familiar with simple type systems that ensure that the code they write is type safe in the context of the language they are writing it in. However, much more sophisticated type systems exist that can be used to ensure specific properties of a program, such as resource usage (Linear Types~\cite{linear}), value predicates (Liquid Types~\cite{DBLP:conf/pldi/RondonKJ08}), or correct following of a distributed protocol (Session Types~\cite{session}).

@@

In this talk, we try to address the challenge of how can multiple type system co-exist in the same programming language, allowing different, valid combinations to be used in the type-checking of a program.

@@

\begin{minipage}{0.4\linewidth}
\begin{lstlisting}[caption={Simple example code},label={lst:code_before}]
printLines lines idx len {
  if len != idx then {
    line = get(lines, idx)
    print(line)
    printLines(lines, idx+1, len)
  }
}

@@

\subsection{Goals}
\label{sec:proposed-approach}

The Goals of the Scholarship was to explore a new verification approach, called \LayeredTypes. The Goal was, that in \LayeredTypes, developers can write programs, but can also define additional type systems as layers. Each layer defines the basic types, and how typechecking happens. Additionally, a layer may depend on another layer if it requires information (such as types or typing contexts) from another layer.

\subsection{Results obtained}
During the duration of the scholarship a prototype compiler and verifier was built in a prototype language. This work layed an important foundation for the subsequent work for the remaining time of the ERASMUS+ Internship in which the implementation was further extended and improved. Two exemplary use-cases were developed showing the potential of the layered approach.

One example of how the framework works can be seen in the accompanging code fragments. In \Cref{lst:code_after} we see a set of annotations that can be added to the base code of \Cref{lst:code_before}. We define three separate layers \texttt{state}, \texttt{types} and \texttt{liquid}. Internally, the framework will build a dependency graph thus allowing to verify properties independently from one another where appropriate.

The full implementation of the developed prototype will be published on GitHub\footnote{\url{https://github.com/LuAbelt/LayeredTypes}}. In addition a submission for a contributed talk was made to the TYPES2023 conference\footnote{\url{https://types2023.webs.upv.es/Index.html}} that is still pending a notification of acceptance or rejection. Additionally it is planned to potentially submit a paper to a relevant conference later in 2023.

\input{../chapters/acknowledgements.tex}

\bibliographystyle{plain}
\bibliography{literature.bib}

>>>>

\usepackage{cleveref}
\usepackage{doc}
\usepackage{tabularx}
% use this if you have a long article and want to create an index
% \usepackage{makeidx}

\lstset{
  frame=none,
  xleftmargin=2pt,
  stepnumber=1,
  numbers=left,
  numbersep=5pt,
  numberstyle=\ttfamily\tiny\color[gray]{0.3},
  belowcaptionskip=\bigskipamount,
  escapeinside={*'}{'*},
  tabsize=2,
  emphstyle={\bf},
  commentstyle=\color{ForestGreen},
  stringstyle=\mdseries\rmfamily,
  showspaces=false,
  keywordstyle=\bfseries\rmfamily,
  columns=flexible,
  basicstyle=\small\sffamily,
  showstringspaces=false,
  morecomment=[l]\%,
}

\lstset{basicstyle=\footnotesize\ttfamily,breaklines=true}
% In order to save space or manage large tables or figures in a
% landcape-like text, you can use the rotating and pdflscape
% packages. Uncomment the desired from the below.
%
% \usepackage{rotating}
% \usepackage{pdflscape}

% Some of our commands for this guide.
%
\authorrunning{Lukas Abelt}

\newcommand{\LayeredTypes}{\textsc{LayeredTypes}}
\newcommand{\Lark}{\textsc{Lark}}
%\makeindex

\makeatletter
\def\input@path{{../chapters}}
%or: \def\input@path{{/path/to/folder/}{/path/to/other/folder/}}
\makeatother
\graphicspath{{../graph}}

%% Front Matter
%%
% Regular title as in the article class.
%
\title{Industry Internship -- Final Report}

@@

% Institutes for affiliations are also joined by \and,
\institute{
   Saarland University,
   Germany\\
   \email{luab00001@stud.uni-saarland.de}
  \and
   LASIGE,
  Faculdade de Ciências da Universidade de Lisboa, Portugal\\
  \email{labelt@lasige.di.fc.ul.pt},
}

\titlerunning{Industrial Internship Report}

@@

%------------------------------------------------------------------------------
\section{About this document}

This document composes the internship report for the industrial internship of Lukas Abelt, carried out from October 2022 until March 2023 at the LASIGE Research Centre in Lisbon, Portugal.

\section{Internship Overview}
This seciton gives an overview of the relevant key data of persons and institutions involved in this internship.

\begin{tabularx}{\textwidth}{rX}
	\textbf{Name of Intern/Student:} & Lukas Abelt\\
	\textbf{Student Contact Information:} & \email{luab00001@stud.uni-saarland.de}\\
	\textbf{Matriculation Number at UdS:} & 7009048\\
	\textbf{Course of Study at UdS:} & Master of Science Computer Science\\
	\textbf{Start of Internship:} & 01.01.2022\\
	\textbf{End of Internship:} & 31.03.2023\\
	\textbf{Hosting Insitution:} & LASIGE Research Unit\footnote{\url{https://www.lasige.pt/}}\newline Reliable Software Systems Research Line\newline Faculdade de Ciências da Universidade de Lisboa\\
	\textbf{Internship Supervisor:} & Alcides Fonseca, Assistant Professor\\
	\textbf{Supervisor Contact Information:} & \email{amfonseca@fc.ul.pt}\\
	\textbf{Internship Title:} & \textit{Type Layers for Gradual Verification}
\end{tabularx}

\section{Internship Description}
This section will give a bread overview of the work that was carried out during the internship. It wstart by providing context to the problem at hand and continue further with the proposed approach and goals of the internship. A technical description of the implemented solution will be provided and finally a short overview of the acquired skills and produced deliverables will be provided.

\subsection{Context}
\label{sec:context}
Most programmers and computer scientists will be familiar with simple type systems that ensure that the code they write is type safe in the context of the language they are writing it in. However, much more sophisticated type systems exist that can be used to ensure specific properties of a program, such as resource usage (Linear Types~\cite{linear}), value predicates (Liquid Types~\cite{DBLP:conf/pldi/RondonKJ08}), or correct following of a distributed protocol (Session Types~\cite{session}).

@@

During this internship, we tried to address the challenge of how can multiple type system co-exist in the same programming language, allowing different, valid combinations to be used in the type-checking of a program.

@@

\subsection{Goals}
\label{sec:proposed-approach}
The Goals of the internship was to explore a new verification approach, which was later decided to be called \LayeredTypes. In \LayeredTypes, developers can write programs, but can also define additional type systems as layers. Each layer defines the basic types, and how typechecking happens. Additionally, a layer may depend on another layer if it requires information (such as types or typing contexts) from another layer.

\subsection{Implementation Details}

We developed a prototype for a layered verification approach called \LayeredTypes. In this prototype, programs can be written in a simple functional language that supports some primitive operations such as variables, function definitions, basic binary operations and \texttt{let} constructs. To enable the use of more complex functionality, it is also possible to define additional functions in python that can be called in this simple language. An example of a program written in that language can be seen in \Cref{lst:code_before}.

The prototype consists of a simple compiler and interpreter for said functional language. The prototype has been developed in Python using the \Lark compiler framework. The core of our compiler is a \textit{layered static verifier}. Within a program, the programer can define annotations to be passed on to one or multiple verification layers. The implementation of each verification layer is provided by the programer in a separate python file with a pre-defined set of functions that will be called by the verifier. Specifically, each layer can provide the following information:

\begin{itemize}
	\item A list of all other layers that it depends on -- That is, all layers that need to be verified successfully before the layer itself is verified
	\item A list of all layers that should be run \textit{after} this layer has been successfully verified
	\item A \texttt{typecheck} function that gets passed the AST of the program and can perform the specific verification for the current layer.
\end{itemize}

The framework will build the dependency graph of all layers defined and used in a program and run them in a topological order so that all dependencies are satisfied. A conceptual overview of the verification framework can be seen in \Cref{fig:framework_overview}

\begin{figure}[ht!]
	\centering
	\includegraphics[width=.8\textwidth]{framework_overview}
	\label{fig:framework_overview}
	\caption{Conceptual Overview of \LayeredTypes}
\end{figure}

To evaluate our work, we also implemented two different use-cases:

In the first Use-Case, we developed a simple program that reads input from a file and processes it line by line. In this use-case we use three different layers. The first layer, \texttt{types} is a simple layer that is used to annotate typing information to variables and functions. The second layer, \texttt{typecheck}, depends on the first one and defines rules on how the program should be type-checked. It then uses the annotations provided by the \texttt{type} layer to check if the program is well-typed. The third layer, \texttt{state}, can be used to define, check and transistion states on variables. In our example we use it to track the state of the file descriptor and ensure that it is closed at the end of the program. The \texttt{state} layer acts independently from the other two layers.

In the second Use-Case we focus on a problem currently present in implementations for \textit{Liquid Types}:

The full implementation of the developed prototype of \LayeredTypes will be published on GitHub\footnote{\url{https://github.com/LuAbelt/LayeredTypes}}. In addition a submission for a contributed talk was made to the TYPES2023 conference\footnote{\url{https://types2023.webs.upv.es/Index.html}} that is still pending a notification of acceptance or rejection. Additionally it is planned to potentially submit a paper to a relevant conference later in 2023.

\input{acknowledgements}

\bibliographystyle{plain}
\bibliography{literature.bib}
\begin{minipage}{0.4\linewidth}
\begin{lstlisting}[caption={Simple example code},label={lst:code_before}]
printLines lines idx len {
  if len != idx then {
    line = get(lines, idx)
    print(line)
    printLines(lines, idx+1, len)
  }
}

@@
